Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-16 appendix.tex
rearranging; cleaning TODOs; felsenstein argument

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:gen-sitepatprob} contains calculations of site split probabilities for the InvFels topology with arbitrary branch lengths and Table~\ref{tab:sitepatprob} site split probabilities for the InvFels topology in Fig.~\ref{fig:farris-fels-top}a.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{1}$       &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{2}$       &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{3}$       &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{12}$     &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{13}$     &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{23}$     &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{123}$   &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:gen-sitepatprob}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x,y,x,y,y\})$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+x^2+y^2+4xy^2+x^2y^2$\\
    $\{1\}$     & $p_{1}$           &$1-x^2+y^2-x^2y^2$\\
    $\{2\}$     & $p_{2}$           &$1+x^2-y^2-x^2y^2$\\
    $\{3\}$     & $p_{3}$           &$1-x^2+y^2-x^2y^2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1+x^2-y^2-x^2y^2$\\
    $\{1,2\}$   & $p_{12}$          &$1-x^2-y^2+x^2y^2$\\
    $\{2,3\}$   & $p_{23}$          &$1-x^2-y^2+x^2y^2$\\
    $\{1,3\}$   & $p_{13}$          &$1+x^2+y^2-4xy^2+x^2y^2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the InvFels tree $\tau_1$ with $t=\{x,y,x,y,y\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{|ll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplit_k$ & $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$\\
\hline
$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
&$\{2\}^*$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
&$\{1,2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\

$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$\\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$\\

$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$\\

$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{2\}^*$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{2\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$\\

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1)$.
There are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
There are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{proof}
We obtain
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau^*, t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau^*, \hat{t})
\]
as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
\[
\shannonDivergence_{\tau^*,t^*}(\tau^*,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
\]
and
\[
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau^*, t).
\]
By straightforward calculus, the function $f(x)=a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$, and so we have a closed form for $\hat{t}$.
Using this estimate, we have the lower bound
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau^*, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau^*,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau^*, \hat{t}).
\]
Figure~\ref{fig:max-anc-state} shows where the lower bound above is greater than all 80 other upper bounds.
\end{proof}

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the divergence term
\[
\shannonDivergence_{\tau^*,t^*}(\tau^*,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau^*,t) \mid_{t=t^*} = 0.
\]
We now show the gradient of the divergence term is a decreasing function on $(0,1)$ in each variable.
We focus on $x_1$.
The divergence proportional to $x_1$ can be written as
\begin{align*}
\shannonDivergence(x_1) \propto \sum_j p_{\siteSplit_j} \log(c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
\end{align*}
where $c_{\siteSplit_j}$ and $a_{\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example,
\[
c_{\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\shannonDivergence'(x_1) := \frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
\end{align*}
and focus on a single term
\[
f_j(x_1) = p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
\]
Regardless of $a_{\siteSplit_j} \lessgtr 0$, this is a decreasing function of $x_1$ provided $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 > 0$, but we see
\[
p_{\siteSplit_j} := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
\]
and $p_{\siteSplit_j} > 0$ so long as $x^*\in(0,1)$ and $y^*\in(0,1)$.
The remaining variables follow similar arguments.

>>>>

We state the Hadamard representation of site split generating probabilities---that is, probabilities of obtaining particular site splits given a tree---following Section 8.6 of \citet{Semple2003-em}.
For each edge $e$ define the edge ``fidelity'' for that edge as
\[
\theta(e) = 1-2p(e)
\]
where $p(e)$ is the probability of a character change along edge $e$.
For an even-sized subset of $Y\subseteq\mathcal{S}$, let the path set $P(Y)$ be the set of edges in the path connecting both elements of $Y$.
For $n$ taxa, the probability of observing site split $A\in\mathcal{P}(\siteSplitSet)$ is
\begin{equation}
\label{eq:hadamard_probability}
p_A = \frac{1}{2^{n-1}} \ \sum_{Y \subseteq \mathcal{S} : |Y| \equiv 0 (\mathrm{mod} \ 2)} \ \left[(-1)^{|Y \cap A|} \ \prod_{e\in P(Y)} \ \theta(e) \right].
\end{equation}
By convention, we set $P(\emptyset)=\emptyset$ and $\prod_{e\in\emptyset} \ \theta(e) = 1$.
For notational convenience, let
\[
p_{\siteSplit_j} := \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,t),
\]
for any site split $\siteSplit_j$.
Table~\ref{tab:gen-sitepatprob} contains calculations of site split probabilities for the trees in Fig.~\ref{fig:farris-fels-top}.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \multicolumn{3}{c}{InvFels tree $\tau=\tau^*$, $t^*=\{x^*,y^*,x^*,y^*,y^*\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{1\}$     & $p_{1}$           &$1-(x^*)^2+(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{2\}$     & $p_{2}$           &$1+(x^*)^2-(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{3\}$     & $p_{3}$           &$1-(x^*)^2+(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1+(x^*)^2-(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{1,2\}$   & $p_{12}$          &$1-(x^*)^2-(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{2,3\}$   & $p_{23}$          &$1-(x^*)^2-(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{1,3\}$   & $p_{13}$          &$1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2$\\
    \hline
    \multicolumn{3}{c}{InvFels tree $\tau=\tau_1$, $t=\{x_1,y_1,x_2,y_2,w\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{1}$           &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{2}$           &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{3}$           &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{12}$          &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{23}$          &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{13}$          &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    \hline
    \multicolumn{3}{c}{Felsenstein tree $\tau=\tau_2$, $t=\{x_1,y_1,x_2,y_2,w\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1y_1 +  x_2y_2 +  w[  x_1 + y_1][  x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{1}$           &$1 - x_1y_1 +  x_2y_2 +  w[ -x_1 + y_1][  x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{2}$           &$1 - x_1y_1 +  x_2y_2 +  w[  x_1 - y_1][  x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{3}$           &$1 + x_1y_1 -  x_2y_2 +  w[  x_1 + y_1][ -x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1 + x_1y_1 -  x_2y_2 +  w[ -x_1 - y_1][ -x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{12}$          &$1 + x_1y_1 +  x_2y_2 +  w[ -x_1 - y_1][  x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{23}$          &$1 - x_1y_1 -  x_2y_2 +  w[  x_1 - y_1][ -x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{13}$          &$1 - x_1y_1 -  x_2y_2 +  w[ -x_1 + y_1][ -x_2 + y_2] + x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the true InvFels tree $\tau^*$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, and on the InvFels tree $\tau_1$ and Felsenstein tree $\tau_2$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
All values multiplied by $1/8$.}
\label{tab:gen-sitepatprob}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\hline
$\siteSplit_j$ & $\ancestralSplit_k$ & $\Pr(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau_1, t)$\\
\hline
$\emptyset$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$             \\
&$\{2\}^*$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$           \\

$\{1\}$    &$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1+w)$             \\

$\{2\}$    &$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$             \\
&$\{2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$               \\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$             \\

$\{3\}$    &$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1+x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1-x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\

$\{1,2,3\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$             \\

$\{1,2\}$  &$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1-w)$               \\
&$\{2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1-w)$               \\
&$\{1,2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1+w)$             \\

$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}$&$(1+x_1)(1+y_1)(1-x_2)(1-y_2)(1-w)$               \\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$             \\

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values multiplied by $1/32$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1)$.
There are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

$\{2,3\}$  &$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{1,2\}$&$(1-x_1)(1+y_1)(1+x_2)(1-y_2)(1+w)$\\

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral states to reduce the number of possible functional forms to consider.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
There are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{proof}
We obtain
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau^*, t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau^*, \hat{t})
\]
as an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} using Gibbs's inequality
\[
\shannonDivergence_{\tau^*,t^*}(\tau^*,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
\]
and
\[
\hat{t} = \argmax_{t} \ \tilde{\ell}_{\tau^*,t^*}(\tau^*, t).
\]
By straightforward calculus, the function $f(x)=a_1\log(1+x)+a_2\log(1-x)$ attains a maximum at $\hat{x} = (a_1-a_2)/(a_1+a_2)$, and so we have a closed form for $\hat{t}$.
Using this estimate, we have the lower bound
\[
\max_{t} \ \ell_{\tau^*,t^*}(\tau^*, t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau^*,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau^*, \hat{t}).
\]
Fig.~\ref{fig:max-anc-state} shows where the lower bound above is greater than all 80 other upper bounds.
\end{proof}

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the divergence term
\[
\shannonDivergence_{\tau^*,t^*}(\tau^*,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau^*,t) \mid_{t=t^*} = 0.
\]
We now show the gradient of the divergence term is a decreasing function on $(0,1)$ in each variable.
We focus on $x_1$.
The divergence proportional to $x_1$ can be written as
\begin{align*}
\shannonDivergence(x_1) \propto \sum_j p_{\siteSplit_j} \log(c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
\end{align*}
where $c_{\siteSplit_j}$ and $a_{\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example,
\[
c_{\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\shannonDivergence'(x_1) := \frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
\end{align*}
and focus on a single term
\[
f_j(x_1) = p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
\]
Regardless of $a_{\siteSplit_j} \lessgtr 0$, this is a decreasing function of $x_1$ provided $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 > 0$, but we see
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
\]
which is strictly greater than zero provided $x_1,y_1,x_2,y_2\in(0,1)$.
The remaining variables follow similar arguments.
